Interactive Curvature Tensor Visualization on Digital Surfaces\thanks{This work has been mainly funded by \textsc{DigitalSnow} ANR-11-BS02-009, \textsc{KIDICO} ANR-2010-BLAN-0205 and \textsc{PRIMES Labex} ANR-11-LABX-0063/ ANR-11-IDEX-0007 research grants.}

### 2015-10-23 latex/article.tex
papier version quasi finale

In \cite{DGCI2013}, authors have defined the 2D digital curvature estimator
$\CurvH{R}$ and the 3D digital mean curvature estimator $\MeanCurvH{R}$ based on
the digital volumetric integral $\VolC(Y,h) \EqDef h^d \MCard(Y)$ with
$\MCard(Y)$ the number of digital points of the intersection between a ball
$\Ball{R/h}{\vx/h}$ of digital radius $R/h$ and $\DigShape$ (we denote such
quantity $\AreaC(Y,h)$ in dimension 2):
%
\begin{Definition}
  %
  Given the Gauss digitization $Z\EqDef\DSh$ of a shape $\Shape \subset \R^2$
  (or $\R^3$ for the 3D mean curvature estimator), {\em digital curvature
  estimators} are defined for any point $\vx \in \R^2$ (or $\R^3$) as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    &\CurvH{R}(Z,\vx,h) \EqDef \frac{3\pi}{2R}
    - \frac{3\AreaC(\Ball{R/h}{\vx/h} \cap
    {Z}, h)}{R^3}\,,\label{eq:dig-curvature-estimator-k}\\
    &\MeanCurvH{R}(Z,\vx,h) \EqDef \frac{8}{3R}
    - \frac{4\VolC(\Ball{R/h}{\vx/h} \cap
    Z, h)}{\pi R^4}\,.\label{eq:dig-curvature-estimator-H}
  \end{align}
  %
  where $\Ball{R/h}{\vx/h}$ is the ball of digital radius $R/h$ centered on
  digital point $(\vx/h) \in Z$.
  %
  % \AreaC(\Ball{R/h}{\vx/h} \cap \Body{\DSh}{h}, h) \EqDef h^2 \MCard(
  % \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ and $\VolC(\Ball{R/h}{\vx/h} \cap
  % \Body{\DSh}{h}, h) \EqDef h^3 \MCard( \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ are
  % the number of digital points in the intersection between $\Body{\DSh}{h}$ and
  % the ball $\Ball{R/h}{\vx/h}$ of radius $R/h$ centered on $\vx/h$, scaled by
  % $h^d$.\todo[inline]{J: Hum... pas top...}
  %
  % Similarly, $\hat{\lambda}_1$
  % and $\hat{\lambda}_2$ are the two greatest eigenvalues of the covariance
  % matrix of the digital points in the intersection between $Z$ and the ball.
  %
\end{Definition}
%
They have proved that estimating the 2D curvature and 3D mean curvature on digital
objets with volumetric integral information have the multigrid convergence
property \cite{DGCI2013}. For short, when the digital objet becomes finer and
finer, \ie when the digitization step $h$ tends to zero, the estimated quantity
on $\Bd{\Body{\DSh}{h}}$ converges (theoretically and experimentally) to the
associated one on $\Bd{\Shape}$.
%
%Since our digital estimators are defined on
%digital shapes, in order to compare results with the exact curvature on
%Euclidean shape, we need to match points on boundary of the digital
%shape and points on the border of the Euclidean one.
%
Setting $R = kh^\frac{1}{3}$ on convex shapes
with at least $C^3$-boundary and bounded curvature, where $k$ is a constant
related to the maximal curvature of the shape, they have:
%
\begin{align}
\forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
\forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
&\CurvH{R}(\DSh, \vxH, h) = \Curv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,,\\
&\MeanCurvH{R}(\DSh, \vxH, h) = \MeanCurv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,.
\end{align}

In \cite{CVIU2014}, authors have also defined the 3D digital principal curvature
estimators $\PrincCurvH{1}{R}$ and $\PrincCurvH{2}{R}$ on $Z\subset\Z^3$ based
on digital moments:
%
\begin{Definition}
  %
  Given the Gauss digitization $Z \EqDef \DSh$ of a shape $\Shape \subset \R^3$,
  {\em 3D digital principal curvature estimators} are defined for any point $\vx
  \in \R^3$ as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    \PrincCurvH{1}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_2 - 3\hat{\lambda}_1) + \frac{8}{5 R} \,,\label{eq:dig-curvature-estimator-k1}\\
    \PrincCurvH{2}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_1 - 3\hat{\lambda}_2) + \frac{8}{5 R} \,,
    \label{eq:dig-curvature-estimator-k2}
  \end{align}
  %
  where $\hat{\lambda}_1$ and $\hat{\lambda}_2$ are the two greatest eigenvalues
  of the covariance matrix of the intersection between the digital shape
  $\DigShape$ and a ball of digital radius $R/h$ centered on digital point
  $\vx/h$.
  %
\end{Definition}
%
The covariance matrix needs to compute digital moments of order $0$, order $1$
and order $2$ (see Equation~20 of \cite{CVIU2014} for more details).
% ($\DMom{000}{h}$), of order $1$ ($\DMom{100}{h}$, $\DMom{010}{h}$ and
% $\DMom{001}{h}$) and of order $2$ ($\DMom{200}{h}$, $\DMom{020}{h}$,
% $\DMom{002}{h}$, $\DMom{110}{h}$, $\DMom{011}{h}$ and $\DMom{101}{h}$, see
% Equation~20 of \cite{CVIU2014} for more details).
They have also proved the convergence of these estimators when setting the ball
radius $h$ in $R = kh^\frac{1}{3}$, where $k$ is a constant related to the
maximal curvature of the shape, on convex shapes with at least $C^3$-boundary
and bounded curvature \cite{CVIU2014}. More formally, they get:
%
\begin{align}
  \forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
  \forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
  &\PrincCurvH{1}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{1}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,,\\
  &\PrincCurvH{2}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{2}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,.
\end{align}

Additionally, eigenvectors associated with $\hat{\lambda}_1$ and
$\hat{\lambda}_2$ of the covariance matrix correspond to principal curvature
directions $\PrincDirH{1}{R}$ and $\PrincDirH{2}{R}$. The smallest eigenvector
corresponds to the normal direction $\NormalDirH{R}$ at $\vx$. Convergence
results can be found in \cite{CVIU2014}.%%% Changer avec le chapitre de Jaco à un moment

It has been shown that the radius of the ball depends on the geometry of the
underlying shape. In \cite{DGCI2014}, authors proposed a parameter-free
estimation of the radius of the ball by analyzing the shape \wrt the local shape
geometry using maximal digital straight segments of the digital boundary. In
\cite{SMI2015}, authors analyzed in scale-space (for a range of radii) these
estimators for a given digital shape. This allows them to detect features of the
shape thanks to the behavior of the estimators on singularities.

@@

In this section, we detail the adaptive isosurface extraction
algorithm.
%The proposed approach is based on an octree representation
%of the input object on which an efficient adaptive triangulation
%of the isosurface is constructed on GPU.
The proposed approch use an octree representation of the input object
on which an adaptive Marching Cube is used to build the isosurface efficiently.
Such approach is motivated by
the fact that the hierarchical representation of the object allows us
to handle large datasets and to locally adapt the level of details
\wrt the geometry or camera position. We first present the
octree representation and then the isosurface extraction from
hierarchical octree cells.

@@

Representing a hierachical structure on GPU is usually challenging since
the GPU is a data parallel component and is unable to handle recursivity.

@@

Efficient spatial tree encoding can be
achieved using pointerless structures such as linear quadtrees or octrees (see
for instance \textsc{Gargantini}~\cite{gargantini1982effective}). This structure
%consists in indexing each cell by a \emph{Morton code}: the code of children
indexes each cell by a \emph{Morton code}: the code of children
cells are defined by the code of the parent suffixed by two bits (in dimension
2, three bits in dimension 3) (see Figure
\ref{fig_quadtree_partitionning}-\emph{left}). A cell's code encodes its
position \wrt its parent cell and its complete path to the tree root.
%Hence, whatever their depth, all cell codes are simply stored in a linear vector on memory.
Hence, the tree is fully represented as a linear vector of its leaves, whatever their depth.
Furthermore, cell operations such as subdivision, merging, neighbors
access can be efficiently implemented using bitwise operations on the Morton
code. In the following, we use the GPU implementation proposed by \textsc{Dupuy}
\textit{et al.}~\cite{dupuy2014quadtrees} that we extended to the third dimension.
By using their approach, Morton codes
and bitwise operations are optimized to match with GPU hardware requirements.

@@

Using this spatial data structure, triangulated mesh can be constructed
from a Marching Cubes approach \cite{lorensen1987marching} (MC for short): the
triangulation is generated from local triangle patches computed on
local cell configurations. Since triangle patches are locally
computed, such approach is fully parallel and easy to implement on GPU
hardwares. However, since adjacent cells may not have the same depth
in the octree, original \textsc{Lorensen} and \textsc{Cline}'s rules need to be
updated (see Figure~\ref{fig_quadtree_partitionning}-\emph{right}). Many authors have addressed this problem both for primal and
dual meshes
\cite{shu1995adaptive,schaefer2004dual,lengyel2010voxel,DBLP:journals/cgf/LewinerMPPL10,DBLP:journals/cvgip/LobelloDD14}.

In the following, we use the extension of MC configurations to handle
adaptive structures proposed by \textsc{Lengyel} \emph{et al.}~\cite{lengyel2010voxel}.
First, this approach constrains the octree
structure in order to make sure that the depth difference between any
two adjacent cells is at most one.
Then, \textsc{Lengyel} \emph{et al.} introduce the concept of transition cells.
Those cells are defined to be inserted between two neighbouring octree cells of different depth.
With specific MC configurations for such transition cells, a crack free triangulation can
be extracted.
%Then, \textsc{Lengyel} \emph{et al.} propose
%specific MC configurations for such transition cells,  leading to a
%crack-free triangulation.

Similarly to original MC algorithm, this approach is perfectly adapted
to GPU hardwares: Given a set of cells (a vector of morton codes),
each triangle patch can be constructed in parallel (done by
the \emph{geometry shader} in the GPU graphic pipeline).

@@

As illustrated in Figure~\ref{fig_quadtree_partitionning}-\emph{middle}, we
propose a viewpoint dependent criterion to decide if a cell needs to be refined:
the closer we are to the camera, the finer the cells are.
%In addition to the
%distance criterion, we define an angular criterion to prevent refining cells
%outside the camera frustrum.
Using a criterion based on the distance between a
cell and the camera is well suited to GPU since it can be evaluated
independently on all cells. Hence, we can efficiently determine if a cell must
be subdivided or not. Figure~\ref{fig_lod_octree} illustrates the level of
details (LoD for short) criterion we use. More precisely, in dimension 2, if
$\alpha$ denotes the viewing angle, an object at distance $d$ from the camera
with diameter greater than $2\cdot d\cdot\tan(\alpha)$ is projected into more
than one pixel (see Figure~\ref{fig_lod_octree}-\emph{left}). Hence, the
distance criterion is based on the ratio (\emph{visibility ratio} in the
following) between the cell diameter $l(c)$ (power of 2 depending on the depth),
and $2\cdot d_c\cdot\tan(\alpha)$, $d_c$ being the distance between the cell
center and the camera. For a given cell $c$, split and merge decision are based
on this visibility ratio:
%
\begin{itemize}
  \item $c$ is split if its children
cells have a visibility ratio greater than  constant $k$\,;
  \item $c$ and its sibling cells are merged if their parent cell $c'$ has a visibility ratio lower than $k$\;
  \item otherwise, the cell $c$ stays for the next frame.
\end{itemize}
%
\begin{figure}[!h]
  \begin{center}
    \begin{overpic}[width=3.5cm]{figs/criterion}
      \put(33,40){$\alpha$}
      \put(50,45){$d_c$}
      \put(95,65){$l(c)$}
    \end{overpic}~~~~~
    \includegraphics[width=3cm]{viewlod2_small}~~~~~
    \includegraphics[width=3cm]{subdivision}
  \end{center}
  \caption{Notations \emph{(left)} and adaptive meshing in dimension 2 using the LoD distance and angular criterion \emph{(middle)}. Based on LoD criterion
    evaluation, the cell buffer is updated between each frame \emph{(right)}.}
  \label{fig_lod_octree}
\end{figure}
%
In addition to the distance criterion, we add an angular term in order
to prevent subdivision of cells far from the camera frustrum (see
Figure~\ref{fig_lod_octree}-\emph{middle}). We do not go into the
details but please note that stay, split and merge decisions of all cells are computed in
parallel from all cell morton codes.

Once decisions have been made, a new set of cells are sent to the mesh
generation step described above. Updating the cell buffer can be
efficiently implemented on GPU hardware as illustrated in
Figure~\ref{fig_lod_octree}-\emph{right}.  Finally, before
constructing the triangulation from remaining cells and transition
cells, geometrical culling is performed in order to skip the triangle
patch construction which are not visible. Figure~\ref{fig_pipeline}
illustrates the overall fully data parallel pipeline.

@@

\paragraph{Hierarchical Decomposition.}  As discussed above, best
results in terms of texture fetch can be obtained by the approach
presented in Fig.~\ref{fig:approx2}-$b$. At each vertex, the
computation algorithm would consist in a recursive decomposition of
the domain such that a cell is subdivided if it does not belong to
$\Ball{R}{\vx}\cap\Shape$. However, on GPU hardware, this approach is
highly inefficient since the hardware optimizes the parallelism only
when the micro-program described in the shader has a predictive
execution flow. Hence, implementing the recursive algorithm (or its
de-recursified version) requires a lot of branches (conditional
\emph{if} tests) in the flow. Instead of implementing this approach,
we have considered a fast hierarchical approximation following
Fig.~\ref{fig:approx2}-$a$: for a given radius $R$, we precompute a
hierachical decomposition of $B_R$. Such decomposition is made of
octree cells at different resolutions. At a given MC vertex $\vx$, the
algorithm becomes simple since we just scan the ball cells and
estimate the area (or other moments) from the mipmap values associated
to this cell. Note that cells in the $B_R$ decomposition may not be
aligned with mipmap cells. However, we can use GPU mipmap
functionalities which interpolates mipmap values if we probe at
non-discrete positions. For a given radius $R$, the expected number of
cells on $B_R$ is in $O(\log{R})$. Implementation details on the
hierarchical octree representation can be found in the supplementary
material (see Section \ref{sec:discussion}).

@@

We first evaluate the curvature estimation on a full-resolution geometry
obtained by disabling the LoD criterion at the mesh generation step.
Figure~\ref{fig:full}-top and middle shows results of curvature tensor
estimation (mean, Gaussian, first and second principal curvatures, first and
second principal directions) on an OctaFlower at level $l_0$ (ground truth).
Figure~\ref{fig:full}-bottom shows mean curvature estimation in real-time on a
dynamic object. For this case, we simply evaluate the implicit expression at
each vertex of a cell to decide if such cell is included or not into
$\Ball{R}{\vx}\cap \Shape$ instead of updating the \emph{mipmap texture} of
densities at each time step. For all illustrations, we use directly normal
vectors computed by algorithm discussed in Section~\ref{sec:curv-tens-estim}.

Then, we compare the approximation made by computing curvature from level $l \ge
l_0$ in Figure~\ref{fig:refine}. We can see that results using approximation
seems to fastly converge to the ground truth results.
Table~\ref{tab:full-res-stat} shows numerical results of $L_\infty$ error (worst
case) and $L_2$ error (mean), as well the number of \emph{mipmap texture}
fetches. We can also see that the number of texel fetch, required to compute the
curvature, reduces drastically when computing approximations. When measuring
results with higher levels, we get consequently much bigger errors. Those level
are thus never used in practice, they are presented here to illustrate how the
refining process converges to the $l_0$ level. We see that levels close to $l_0$
give few errors than $l_0$, and should be considered for exploring our datasets
in real time, changing parameters on-the-fly and visualizing dynamic datasets.

We also compare them with the hierarchical algorithm (as discussed in
Section~\ref{sec:hierarchical}). This method introduce a higher error than
$l_0$. This is due to the precomputation of the subdivision that no longer
ensures to fetch data at the center of a mipmap cell. Then, the GPU performes an
interpolation to reduce the bias, but it doesn't remove it.

>>>>

In \cite{DGCI2013}, authors have defined the 2D digital curvature estimator
$\CurvH{R}$ and the 3D digital mean curvature estimator $\MeanCurvH{R}$ based on
the digital volumetric integral $\VolC(Y,h) \EqDef h^d \MCard(Y)$ with
$\MCard(Y)$, the number of digital points of the intersection between a ball
$\Ball{R/h}{\vx/h}$ of digital radius $R/h$ and $\DigShape$ (we denote such
quantity $\AreaC(Y,h)$ in dimension 2):
%
\begin{Definition}
  %
  Given the Gauss digitization $Z\EqDef\DSh$ of a shape $\Shape \subset \R^2$
  (or $\R^3$ for the 3D mean curvature estimator), {\em digital curvature
  estimators} are defined for any point $\vx \in \R^2$ (or $\R^3$) as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    &\CurvH{R}(Z,\vx,h) \EqDef \frac{3\pi}{2R}
    - \frac{3\AreaC(\Ball{R/h}{\vx/h} \cap
    {Z}, h)}{R^3}\,,\label{eq:dig-curvature-estimator-k}\\
    &\MeanCurvH{R}(Z,\vx,h) \EqDef \frac{8}{3R}
    - \frac{4\VolC(\Ball{R/h}{\vx/h} \cap
    Z, h)}{\pi R^4}\,.\label{eq:dig-curvature-estimator-H}
  \end{align}
  %
  where $\Ball{R/h}{\vx/h}$ is the ball of digital radius $R/h$ centered on
  digital point $(\vx/h) \in Z$.
  %
  % \AreaC(\Ball{R/h}{\vx/h} \cap \Body{\DSh}{h}, h) \EqDef h^2 \MCard(
  % \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ and $\VolC(\Ball{R/h}{\vx/h} \cap
  % \Body{\DSh}{h}, h) \EqDef h^3 \MCard( \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ are
  % the number of digital points in the intersection between $\Body{\DSh}{h}$ and
  % the ball $\Ball{R/h}{\vx/h}$ of radius $R/h$ centered on $\vx/h$, scaled by
  % $h^d$.\todo[inline]{J: Hum... pas top...}
  %
  % Similarly, $\hat{\lambda}_1$
  % and $\hat{\lambda}_2$ are the two greatest eigenvalues of the covariance
  % matrix of the digital points in the intersection between $Z$ and the ball.
  %
\end{Definition}
%
They have proved that estimating both 2D curvature and 3D mean curvature on digital
objets with volumetric integral information have the multigrid convergence
property \cite{DGCI2013}. For short, when the digital objet becomes finer and
finer, \ie when the digitization step $h$ tends to zero, the estimated quantity
on $\Bd{\Body{\DSh}{h}}$ converges (theoretically and experimentally) to the
associated one on $\Bd{\Shape}$.
%
%Since our digital estimators are defined on
%digital shapes, in order to compare results with the exact curvature on
%Euclidean shape, we need to match points on boundary of the digital
%shape and points on the border of the Euclidean one.
%
Setting $R = kh^\frac{1}{3}$ on convex shapes
with at least $C^3$-boundary and bounded curvature, where $k$ is a constant
related to the maximal curvature of the shape, they have:
%
\begin{align}
\forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
\forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
&\CurvH{R}(\DSh, \vxH, h) = \Curv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,,\\
&\MeanCurvH{R}(\DSh, \vxH, h) = \MeanCurv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,.
\end{align}

In \cite{CVIU2014}, authors have also defined 3D digital principal curvature
estimators $\PrincCurvH{1}{R}$ and $\PrincCurvH{2}{R}$ on $Z\subset\Z^3$ based
on digital moments:
%
\begin{Definition}
  %
  Given the Gauss digitization $Z \EqDef \DSh$ of a shape $\Shape \subset \R^3$,
  {\em 3D digital principal curvature estimators} are defined for any point $\vx
  \in \R^3$ as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    \PrincCurvH{1}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_2 - 3\hat{\lambda}_1) + \frac{8}{5 R} \,,\label{eq:dig-curvature-estimator-k1}\\
    \PrincCurvH{2}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_1 - 3\hat{\lambda}_2) + \frac{8}{5 R} \,,
    \label{eq:dig-curvature-estimator-k2}
  \end{align}
  %
  where $\hat{\lambda}_1$ and $\hat{\lambda}_2$ are the two greatest eigenvalues
  of the covariance matrix of the intersection between the digital shape
  $\DigShape$ and a ball of digital radius $R/h$ centered on digital point
  $\vx/h$.
  %
\end{Definition}
%
The covariance matrix needs to compute digital moments of order $0$, order $1$
and order $2$ (see Equation~20 of \cite{CVIU2014} for more details).
% ($\DMom{000}{h}$), of order $1$ ($\DMom{100}{h}$, $\DMom{010}{h}$ and
% $\DMom{001}{h}$) and of order $2$ ($\DMom{200}{h}$, $\DMom{020}{h}$,
% $\DMom{002}{h}$, $\DMom{110}{h}$, $\DMom{011}{h}$ and $\DMom{101}{h}$, see
% Equation~20 of \cite{CVIU2014} for more details).
They have also proved the convergence of these estimators when setting the ball
radius $h$ in $R = kh^\frac{1}{3}$, where $k$ is a constant related to the
maximal curvature of the shape, on convex shapes with at least $C^3$-boundary
and bounded curvature \cite{CVIU2014}. More formally, they get:
%
\begin{align}
  \forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
  \forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
  &\PrincCurvH{1}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{1}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,,\\
  &\PrincCurvH{2}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{2}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,.
\end{align}

Additionally, eigenvectors associated to $\hat{\lambda}_1$ and
$\hat{\lambda}_2$ of the covariance matrix correspond to principal curvature
directions $\PrincDirH{1}{R}$ and $\PrincDirH{2}{R}$. The smallest eigenvector
corresponds to the normal direction $\NormalDirH{R}$ at $\vx$. Convergence
results can be found in \cite{ChapterIICurvature}.%%% Changer avec le chapitre de Jaco à un moment

It has been shown that the radius of the ball depends on the geometry of the
underlying shape. In \cite{DGCI2014}, authors proposed a parameter-free
estimation of the radius of the ball by analyzing the shape \wrt the local shape
geometry using maximal digital straight segments of the digital boundary. In
\cite{SMI2015}, authors analyzed in scale-space (for a range of radii) these
estimators for a given digital shape. This allows them to detect features of the
shape thanks to the behavior of estimators on singularities.

@@

In this section, we detail the adaptive isosurface extraction
algorithm.
%The proposed approach is based on an octree representation
%of the input object on which an efficient adaptive triangulation
%of the isosurface is constructed on GPU.
The proposed approch uses an octree representation of the input object
on which an adaptive Marching Cube builds the isosurface efficiently.
Such 
%approach is motivated by the fact that the 
hierarchical representation of the object allows us
to handle large datasets and to locally adapt the level of details
\wrt the geometry or camera position. We first present the
octree representation and then the isosurface extraction.
% from hierarchical octree cells.

@@

Representing a hierachical structure on GPU is usually challenging since
such a data parallel component is unable to handle recursivity.

@@

Efficient spatial tree encoding can be
achieved using pointerless structures such as linear quadtrees or octrees 
%(seefor instance
 \textsc{Gargantini}~\cite{gargantini1982effective}. This structure
%consists in indexing each cell by a \emph{Morton code}: the code of children
indexes each cell by a \emph{Morton code}: the code of children
cells are defined by the code of the parent suffixed by two bits (in dimension
2, three bits in dimension 3) (see Figure
\ref{fig_quadtree_partitionning}-\emph{left}). A cell's code encodes its
position \wrt its parent cell and its complete path to the tree root.
%Hence, whatever their depth, all cell codes are simply stored in a linear vector on memory.
Hence, the tree is fully represented as a linear vector of its leaves.%, whatever their depth.
Furthermore, cell operations such as subdivision, merging %, neighbors access 
can be efficiently implemented using bitwise operations on the Morton
code. In the following, we use the GPU 
friendly 
implementation proposed by \textsc{Dupuy}
\textit{et al.}~\cite{dupuy2014quadtrees}.% that we extended to the third dimension.
%By using their approach, Morton codes
%and bitwise operations are optimized to match with GPU hardware requirements.

@@

Using this spatial data structure, a triangulated mesh can be constructed
%from a Marching Cubes approach \cite{lorensen1987marching} (MC for short): the
using Marching Cubes\cite{lorensen1987marching} (MC for short): the
triangulation is generated from local triangle patches computed on
local cell configurations. 
%Since triangle patches are locally computed,
Such approach is fully parallel and easy to implement on the GPU.
However, since adjacent cells may not have the same depth
in the octree, original \textsc{Lorensen} and \textsc{Cline}'s rules need to be
updated (see Figure~\ref{fig_quadtree_partitionning}-\emph{right}). 
Many authors have addressed this problem both for primal and
dual meshes
\cite{shu1995adaptive,schaefer2004dual,lengyel2010voxel,DBLP:journals/cgf/LewinerMPPL10,DBLP:journals/cvgip/LobelloDD14}.

In the following, we use the 
%extension  of MC configurations to handle adaptive structures 
algorithm
proposed by \textsc{Lengyel} \emph{et al.}~\cite{lengyel2010voxel}.
First, this approach constrains the octree
structure 
%in order 
to make sure that the depth difference between any
two adjacent cells is at most one.
Then, \textsc{Lengyel} \emph{et al.} introduce the concept of transition cells.
Those cells are defined to be inserted between two neighboring octree cells of different depth.
With specific MC configurations for their triangulation, a crack free mesh can
be extracted.
%Then, \textsc{Lengyel} \emph{et al.} propose
%specific MC configurations for such transition cells,  leading to a
%crack-free triangulation.

Similarly to original MC algorithm, this approach is well suited%perfectly adapted
to the GPU: Given a set of cells (a vector of morton codes),
each triangle patch can be extracted in parallel for both the regular cell and the transition cells.
%(done by the \emph{geometry shader} in the GPU graphic pipeline).

@@

As illustrated in Figure~\ref{fig_quadtree_partitionning}-\emph{middle}, we
%propose a viewpoint dependent criterion to decide if a cell needs to be refined:
propose a viewpoint dependent criterion to decide if a cell needs to be refined:
the closer we are to the camera, the finer the cells are.
%In addition to the
%distance criterion, we define an angular criterion to prevent refining cells
%outside the camera frustrum.
Such a criterion is also well suited to GPU since it can be evaluated
independently on all cells. 
%Hence, we can efficiently determine if a cell must be subdivided or not. 
Figure~\ref{fig_lod_octree} illustrates our the level of
details (LoD for short) criterion% we use. More precisely, 
In dimension 2, if $\alpha$ denotes the viewing angle, an object at a distance $d$ from the camera
%with diameter greater than $2\cdot d\cdot\tan(\alpha)$ is projected into more
has a projected size on screen of $2\cdot d\cdot\tan(\alpha)$.
%than one pixel 
(see Figure~\ref{fig_lod_octree}-\emph{left}). 
%Hence, the
Our distance criterion is based on the ratio (\emph{visibility ratio} in the
following) between the cell diameter $l(c)$ (power of 2 depending on the depth),
and its projected size. %$2\cdot d_c\cdot\tan(\alpha)$, $d_c$ being the distance between the cell
%center and the camera. 
For a given cell $c$, split and merge decision are based
on this visibility ratio:
%
\begin{itemize}
  \item $c$ is split if its children
cells have a visibility ratio greater than  constant $k$\,;
  \item $c$ and its sibling cells are merged if their parent cell $c'$ has a visibility ratio lower than $k$\;
  \item otherwise, the cell $c$ stays for the next frame.
\end{itemize}
%
\begin{figure}[!h]
  \begin{center}
    \begin{overpic}[width=3.5cm]{figs/criterion}
      \put(33,40){$\alpha$}
      \put(50,45){$d_c$}
      \put(95,65){$l(c)$}
    \end{overpic}~~~~~
    \includegraphics[width=3cm]{viewlod2_small}~~~~~
    %\includegraphics[width=3cm]{subdivision}
  \end{center}
  \caption{Notations \emph{(left)} and adaptive meshing in dimension 2 using the LoD distance and angular criterion \emph{(right)}. 
  %Based on LoD criterion evaluation, the cell buffer is updated between each frame \emph{(right)}.
  }
  \label{fig_lod_octree}
\end{figure}
%
%In addition to the distance criterion, we add an angular term in order
%to prevent subdivision of cells far from the camera frustrum (see
%Figure~\ref{fig_lod_octree}-\emph{middle}). We do not go into the
%details but please note that stay, split and merge decisions of all cells are computed in
%parallel from all cell morton codes.
Using such a criterion, split and merge decisions are computed in%of all cells are computed in
parallel from all cell's morton codes.

Once decisions have been made, a new set of cells is sent to the mesh
generation step described above. 
%Updating the cell buffer can be
%efficiently implemented on GPU hardware as illustrated in
%Figure~\ref{fig_lod_octree}-\emph{right}.  
Finally, before
constructing the triangulation from remaining cells and transition
cells, geometrical culling is performed in order to skip the triangle
patch construction for cells that are not visible. Figure~\ref{fig_pipeline}
illustrates the overall fully data parallel pipeline.

@@

\paragraph{Hierarchical Decomposition.}  As discussed above, best
results in terms of texture fetch can be obtained by the approach
presented in Fig.~\ref{fig:approx2}-$b$. At each vertex, the
computation algorithm would consist in a recursive decomposition of
the domain such that a cell is subdivided if it does not belong to
$\Ball{R}{\vx}\cap\Shape$. However, on the GPU %hardware, 
this approach is
highly inefficient since the hardware optimizes the parallelism only
when the micro-program described in the shader has a predictive
execution flow. Hence, implementing the recursive algorithm (or its
de-recursified version) requires a lot of branches (conditional
\emph{if} tests) in the flow. Instead of implementing this approach,
we have considered a fast hierarchical approximation following
Fig.~\ref{fig:approx2}-$a$: for a given radius $R$, we precompute a
hierachical decomposition of $B_R$. Such decomposition is made of
octree cells at different resolutions. At a given MC vertex $\vx$, the
algorithm becomes simple since we just scan the ball cells and
estimate the area (or other moments) from the mipmap values associated
to this cell. Note that cells in the $B_R$ decomposition may not be
aligned with mipmap cells. However, we can use GPU %mipmap
functionalities which interpolates mipmap values if we probe at
non-discrete positions. For a given radius $R$, the expected number of
cells on $B_R$ is in $O(\log{R})$. Implementation details on the
hierarchical octree representation can be found in the supplementary
material (see Section \ref{sec:discussion}).

@@

We first evaluate the curvature estimation on a full-resolution geometry
obtained by disabling the LoD criterion at the mesh generation step.
Figure~\ref{fig:full}-top shows results of curvature tensor
estimation (mean, Gaussian, first and
second principal directions) on an OctaFlower at level $l_0$ (ground truth).
Figure~\ref{fig:full}-bottom shows mean curvature estimation in real-time on a
dynamic object. For this case, we simply evaluate the implicit expression at
each vertex of a cell to decide if such cell is included or not into
$\Ball{R}{\vx}\cap \Shape$ instead of updating the \emph{mipmap texture} of
densities at each time step. For all illustrations, we use directly normal
vectors computed by algorithm discussed in Section~\ref{sec:curv-tens-estim}.

Then, we compare the approximation made by computing curvature from level $l \ge
l_0$ in Figure~\ref{fig:refine}. We can see that results using approximation
seems to fastly converge to the ground truth results.
Table~\ref{tab:full-res-stat} shows numerical results of $L_\infty$ error (worst
case) and $L_2$ error (mean), as well the number of \emph{mipmap texture}
fetches. We can also see that the number of texel fetch, required to compute the
curvature, reduces drastically when computing approximations. However, at higher levels, 
we get consequently much bigger errors. Those levels
are thus never used in practice, they are presented here to illustrate how the
refining process converges to the $l_0$ level. We see that levels close to $l_0$
lead very few error% than $l_0$, 
and should be considered for exploring our datasets
in real-time, changing parameters on-the-fly and visualizing dynamic datasets.

We also compare them with the hierarchical algorithm (as discussed in
Section~\ref{sec:hierarchical}). This method introduces a higher error when compared with
$l_0$. This is due to the precomputation of the subdivision that no longer
ensures to fetch data at the center of a mipmap cell. 
The GPU's interpolation reduces the bias, but it doesn't remove it.

%Bear also in mind that those results were measured on a full resolution
%triangulation. Since we compute the curvature of each vertices of the final
%mesh, a highly triangulated object will be much slower to process. We therefore
%use in practice an adaptive triangulation of the object, refining the mesh when
%it is closer from the camera and thus reducing the number of vertices to
%compute.

Those results were measured on a full resolution triangulation. 
However, since we compute the curvature of each vertices of the final
mesh, a highly triangulated object will be much slower to process.

When dealing with large datasets, we can not expect real-time curvature tensor
estimation due to the huge amount of data to process. By computing a
dynamically refined approximate curvature tensor estimation joined with an
adaptive triangulation (as discussed in Section~\ref{sec:hierarchical}), we
manage to maintain a real-time framerate by giving control over the amount of
data to process at each frame. In Figure~\ref{fig:timings}, we compare timings
(in logarithmic scale) for a triangulation that is dynamically refining according
to its distance to the camera. We measured those timings using the regular and
the hierarchical algorithm.

%We can first note that the required time for an object to compute ground truth
%curvature is usually as high as the time required to extract the geometry and
%compute all the above levels. Using approximations is thus mandatory, since for
%large objects such as the XYZ-Dragon ($510^3$) with big radii the approximate
%computation is faster than the $l_0$ while results are enough close to the
%$l_0$. For example, with $R=20$, the approximated visualization at $l_2$ takes
%only 300ms while the $l_0$ visualization takes more than 4 seconds.
